---
title: "Class 08: Breast Cancer Mini Project"
author: "Nicole Lin (PID: A17826971)"
format: pdf
---

## Background

In today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data.

## Data import 

The data is in CSV format:

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
wisc.df
```

wee peak at the data
```{r}
head(wisc.df, 3)
```

> Q. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum( wisc.df$diagnosis == "M" )
```
```{r}
table( wisc.df$diagnosis )
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep( "_mean", colnames(wisc.df)))
```

We need to remove the `diagnosis` column before we do any further analysis of this data set - we don't want to pass this to PCA etc. We will save it as a separate wee vector that we can use later to compare our findings to those of experts.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()`. We will use the optional argument `scale=TRUE` here as the data columns/features/dimensions are on very different scales in the original data set.

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
```

```{r}
attributes(wisc.pr)
```

```{r}
library(ggplot2)
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp( wisc.data, scale=T )
```

```{r}
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
 
7


```{r}
biplot(wisc.pr)
```
> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This is very messy and difficult to understand because all the labels and names overlap each other.

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```
The y-axis is different from the PC1 vs PC2 graph becuase PC3 replaces PC2 in this graph.There is more overlap between the benign and malignant data points.


```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- wisc.pr$sdev^2/sum(wisc.pr$sdev^2)
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?









The goal of this section is to do hierarchical clustering of the original data to see if there is any obvious grouping into malignant and benign clusters.

In short, these results are not good!

First we will scale our `wisc.data` then




> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

height at around 18-19.

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

probably ward.D2 because it creates groups that minimizes variance in clusters so it's easier to tell where the merging is happening.


## Combining methods

The idea here is that I can take my new variables (i.e. the scores on the PCs `wisc.pr$x`) that are better descriptors of the data set than the original features (i.e. the 30 columns in `wisc.datat`) and use the as a basis for clustering. 

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method="ward.D2")
plot(wisc.pr.hclust)
```


```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```


```{r}
table(diagnosis)
```

I can now run `table()` with both my clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```
Our cluster "1" has 179 "M" diagnosis
Our cluster "2" has 333 "B" diagnosis

179 TP (true positive)
24 FP (false positive)
333 TN (true negative)
33 FN (false negative)

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)
```{r}
333/(333+24)
```

>Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

The hierarchical clustering without PCA somewhat separates the diagnoses, but there is some overlap between clusters. It captures the general pattern, but it doesn’t perfectly match the true labels.

## Prediction

We can use our PCA model for prediction of new unseen cases:

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

Should follow up on 1 becuase they seem to have malignant cells.




